Collective excitations in competing phases in two and three dimensions

### 2024-03-26 paper/main.tex
conclusion

In Fig.\ \ref{fig:afm_cdw_peaks_details}, we revisit the data of panels (b) and (c) of the previous figure
on a double logarithmic scale.
We restrict the analysis to $\spectral{l.AFM}$ in the CDW phase as the swapped case of $\spectral{CDW}$ in the 
AFM phase shows the same behavior.
We consider $U$ relative to the value $U_0$ at which the peak in $\spectral{l.AFM}$ ceases to exist.
\red{Stimmt die Formulierung?} \blue{Ja.}
This occurs at $U_0 = 3.372t$ on the square lattice and at $U_0 = 4.479t$ on the simple cubic lattice.
The peak positions $\omega_0$ in panels (a) behave quadratically as they merge with the 
two-particle continuum, i.e., 
$\omega_0 \propto (U-U_0)^2$ while the weights $W_0$ in panel (b) decrease linearly, i.e., 
$W_0 \propto (U-U_0)$ \red{Stimmt das Vorzeichen?}. This behavior is generic for the peaks of bound states merging with
their continua at vanishing binding energy \cite{uhrig96b,uhrig96be,zhito13}.

@@

We have conducted an analysis of the mean-field phase diagram of the extended Hubbard model in 2D and 3D. 
Specifically, both systems provide a similar structure of $s$-wave SC, CDW, and AFM.
Furthermore, we were able to identify an area in which our analysis is incomplete.
In 2D, this region coincides with a phase-separated state that has already been observed in recent research \cite{Linner23}.

We have also developed a novel method to obtain Fourier-transformed Green's functions and analyzed various such functions. 
We have found collective excitations within the spectral functions $\spectral{Higgs}$, $\spectral{Phase}$, $\spectral{CDW}$, $\spectral{l.AFM}$, and $\spectral{t.AFM}$ 
and examined their behavior as the system approaches phase transitions. 
In particular, we identified the peak and singularity found in the first two as the well-known phase and amplitude modes in neutral superconductors 
and thereby provided a technique to obtain them starting from a microscopic description.

We observed the behavior of the peak in the SC-related spectral functions, \textcolor{\markEdited}{corresponding to a cooperon}.
It moves towards 0 energy and attains more weight as the system moves from the CDW phase towards the SC phase. 
The peak's relative position to the two-particle continuum and its weight plateau as the system moves away from the phase transition.

In the SC phase close to the CDW phase, the peak in $\spectral{CDW}$, \textcolor{\markEdited}{corresponding to an exciton}, is located at a finite energy and and has a finite weight. 
\textcolor{\markEdited}{It behaves in the same manner as the previously discussed peak}.

Around the AFM-CDW phase transition, both $\spectral{l.AFM}$ and $\spectral{CDW}$ feature a peak. 
\textcolor{\markEdited}{The former corresponds to a longitudinal magnon, while the latter represents an exciton.}
The peak of the former in the AFM phase lies at lower energies than the latter one and vice-versa. 
Moving away from the phase transition, the higher-lying peak moves into the two-particle continuum. 
The peak position relative to the two-particle continuum approaches $0$ quadratically with $U$, and its weight approaches $0$ linearly.

Furthermore, we have found a peak in $\spectral{CDW}$ that appears at large values of $V$, 
which is most likely the same as the one found close to the CDW-AFM phase transition. 
\textcolor{\markEdited}{It is also reasonable to assume, that this peak is always located outside of the two-particle continuum, even for small values of $V$.
In this case, however, it would lie very close to the continuum, which we cannot resolve due to the limited accuracy of our numerics.
Therefore, it appears as a singularity at $2\Delta$.}

@@

Some possible ideas for future research include incorporating phase-separated states into the mean-field theory.
This would enable us to peer deeper into the negative $V$ regime of the phase diagrams and provide more analysis of the collective excitations that occur therein.

\textcolor{\markEdited}{Specifically for the SC-related quantities, one can fathom including a coupling of the electronic system to the electromagnetic field.}
This should move the position of the phase mode close to the plasma frequency as explained by the Anderson-Higgs mechanism.

Additionally, one could describe multi-band systems with even richer phases and collective excitations.

%==============================================================================
% Acknowledgments 
%==============================================================================
\begin{acknowledgments} 
    \textcolor{\markEdited}{This research was partially funded by the MERCUR Kooperation.
    We would like to thank the members of our group for various useful discussions.
    [Oder hier lieber alle mal beim Namen nennen? Ich hab mich ja mit allen mal unterhalten]
    [Götz Seibold zu danken klingt meines Erachtens auch sinnig, mit dem hast du ja viel gesprochen.
    Lido habe ich nicht verwendet.]}
\end{acknowledgments}

>>>>

In Fig.\ \ref{fig:afm_cdw_peaks_details}, we revisit the data of panels (b) and (c) of the previous figure
on a double logarithmic scale.
We restrict the analysis to $\spectral{l.AFM}$ in the CDW phase as the swapped case of $\spectral{CDW}$ in the 
AFM phase shows the same behavior.
We consider $U$ relative to the value $U_0$ at which the peak in $\spectral{l.AFM}$ ceases to exist.
\red{Stimmt die Formulierung?} \blue{Ja.}
This occurs at $U_0 = 3.372t$ on the square lattice and at $U_0 = 4.479t$ on the simple cubic lattice.
The peak positions $\omega_0$ in panels (a) behave quadratically as they merge with the 
two-particle continuum, i.e., 
$\omega_0 \propto (U-U_0)^2$ while the weights $W_0$ in panel (b) decrease linearly, i.e., 
$W_0 \propto (U-U_0)$ \red{Stimmt das Vorzeichen?} \blue{Ja.}. This behavior is generic for the peaks of bound states merging with
their continua at vanishing binding energy \cite{uhrig96b,uhrig96be,zhito13}. \blue{Das wusste ich tatsächlich nicht.}

@@

%%%
% competing phases, collective modes, softening, weights, shift of energies
% controlled, powerful formalism on top of mean-field Hamiltonian,
%but dynamics well beyond!, can be extended to diagrammatics.

The objective of this article was to study collective excitations in competing phases.
We wanted to understand how collective excitations behave in the vicinity of phase transitions
and how they signal these transitions, for example by softening or by losing spectral weight.
To this end, we developed a general and versatile formalism based on iterated equations of motion.
We could show that the formalism reproduces all rigorously known properties such as
the existence of Goldstone bosons in case of broken continuous symmetries and the absence thereof
if the broken symmetry is discrete.

We conducted an analysis of the mean-field phase diagram of the extended Hubbard model in 
two (2D) and three dimensions (3D). The results in both dimensions are qualitatively very similar.
Specifically, we studied the competing phases 
of $s$-wave superconducting order (SC), alternating charge density wave order
(CDW), and N\'eel antiferromagnetism (AFM).
Furthermore, we were able to identify regions in which our analysis is incomplete
in the sense that the assumed phases turn out to be unstable, but it is not obvious which
phases are the stable ones. Since we observed this phenomenon for strongly attractive interactions
we conjecture that the instability indicates the occurrence of phase separation in 2D and 3D. 
For the square lattice, recent findings at finite temperature also pointed in this direction \cite{Linner23}.

We developed a method to obtain Fourier-transformed Green's functions
based on iterated equations of motion and analyzed the most relevant
of them. We identified the signatures of collective excitations in the spectral functions 
$\spectral{Higgs}$, $\spectral{Phase}$, $\spectral{CDW}$, $\spectral{l.AFM}$, and $\spectral{t.AFM}$ 
which are designed to excite the Higgs mode, the phase mode, a staggered density mode, a longitudinal, and a transversal magnon, respectively. 
The focus was the analysis of their behavior as the system approaches and crosses phase transitions.
In particular, we identified the peak and the singularity occurring in the first two 
spectral functions related to superconductivity
to be the well-known amplitude and phase modes in neutral superconductors 
and thereby provided a technique to obtain them starting from a microscopic description.
Outside an SC phase, both spectral functions are identical while in an SC phase,
the phase mode is massless while the Higgs mode is at the lower edge of the continuum.
So, the phase mode manifests as the derivative of the $\delta$ function at zero energy
complying with the asymmetry of the spectral function.

We observed the behavior of the peak in the SC-related spectral functions corresponding to a cooperon in various phases. 
It shifts towards zero energy and acquires more and more weight as the system moves from the CDW phase toward the SC phase. 
The peak's relative position to the two-particle continuum and its weight saturate 
as the system moves away from the phase transition.
In the SC phase close to the CDW phase, the peak in $\spectral{CDW}$ 
is located at a finite energy and and has finite weight. 
It must represent a bound state because it is located below the two-particle continuum. 
This state is induced by the application of an electron creation and an electron annihilation without effect of its spin content so that we interpret it as an $S=0$ exciton.
It behaves similarly to the SC-related peak in non-SC phases. \blue{Explizit in der CDW Phase, nicht?}

In the vicinity of the AFM-CDW phase transition, the spectral densities
$\spectral{CDW}$ and $\spectral{l.AFM}$ feature a peak. 
The former is the aforementioned exciton while the latter is a longitudinal magnon. 
Of course, this can also be seen as an exciton being a bound
state of an electron and a hole, yet with spin content $S=1$ and $S^z=0$. 
The longitudinal magnon lies at lower energy in the AFM phase than the CDW exciton. 
In the CDW phases, the two peaks swap their relative energy positions. 
Moving away from the phase transition, the higher-lying peak merges with the two-particle continuum. 
The peak position relative to the two-particle continuum approaches $0$ 
quadratically in $\Delta U$ while its weight approaches $0$ linearly with $\Delta U$ where
$\Delta U$ measures the distance to the interaction \blue{strength?} where the merging takes place.

Furthermore, we found the exciton peak in $\spectral{CDW}$ also at large values of $V$ in the CDW phase, 
evolving smoothly from the CDW-AFM phase transition. \red{Das kannst Du vllt. noch stichprobenartig pruefen.}
We assume that this peak is always located outside of the two-particle continuum, even for small values of $V$.
In this case, however, it can lie very close to the continuum, so that we cannot resolve it due to the limited accuracy of our numerics.
So it seems to appear as a singularity at $2\Delta$ for small values of $V$.
\red{Stimmst Du dieser Darstellung zu?} \blue{Ja.}

In the AFM phase, the spectral functions $\spectral{l.AMF}$ and $\spectral{t.AFM}$ are distinct.
While the longitudinal magnon continues to manifest as a peak at relatively high energies
the transversal magnon appears as a peak at zero energy within our numerical accuracy. Due to
the asymmetry of the spectral, its peak is the derivative of a $\delta$ function.
It is an advantageous feature of the adopted approach that the Goldstone theorem is 
automatically fulfilled. Both, phase mode and transversal magnon are located at zero energy
if the corresponding symmetry is broken, i.e., the U(1) symmetry and the $O(3)$ symmetry,
respectively. \blue{Mathmode?}

\red{Haben wir jetzt alle wesentlichen Beobachtungen erwaehnt?} \blue{Ja.}

@@

Future directions of research comprise incorporating the investigation of possible phase-separated states. 
This would enable us to delve deeper into the negative $V$ regime 
of the phase diagrams and to analyze its collective excitations. Anticipating the occurrence
of regions of high and low density of electrons, a necessary prerequisite is to investigate doped systems.

Another intriguing extension is to extend the study from the $\Gamma$ point to the full Brillouin zone.
This implies studying the same phases with operators at finite wave vectors beyond the nesting vector. 
Conceptually, this extension is straightforward. But it spoils the efficient approach
to discretize the $\gamma$ space since the dependence on two wave vectors will no longer
allow us to reduce all wave vector dependence to a single dependence on $\gamma$. \blue{Das verwirrt mich.}

Additionally, one can fathom including a coupling of the electronic system to the electromagnetic fields
so that we no longer study a neutral, but a charged superconducting system. One would expect
to capture the important shift of the phase mode to finite energies close to the plasma frequency 
as explained by the Anderson-Higgs mechanism.

Lastly, one could deal with multi-band systems with even richer phase
diagrams and corresponding collective excitations.

%==============================================================================
% Acknowledgments 
%==============================================================================
\begin{acknowledgments} 
	This research was partially funded by the MERCUR Kooperation
		in project KO-2021-0027. We acknowledge very helpful discussions with
		Anna B\"ohmer, J\"org B\"unemann, Ilya Eremin, G\"otz Seibold, Joachim Stolze, and Zhe Wang
		as well as with all members of our research group.
\end{acknowledgments}
